---
title: "**P4 | RNA-seq analysis: Bladder Cancer**<br><font color='#A8A8A8' size='5'>Finding differentiantly expressed genes in bladder cancer</font>"
author: "Current Topics in Bioinformatics || Group 5"
date: "17 January 2022"
output:
  html_document:
    theme: yeti
    css: https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.11.2/css/all.css
    self_contained: yes
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", fig.width = 6, fig.height = 4)
library(SummarizedExperiment)
library(edgeR)
library(DESeq2)
library(tweeDEseq)
library(tweeDEseqCountData)
library(GOstats)
library(annotate)
library(org.Hs.eg.db)
library(biomaRt)
library(ggplot2)
library(ggrepel)
library(devtools)
library(MetBrewer)
```

<style>
  @import url(https://fonts.googleapis.com/css?family=Fira+Sans:300,300i,400,400i,500,500i,700,700i);
  @import url(https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css);
  @import url("https://use.fontawesome.com/releases/v5.10.1/css/all.css");
  body {
  text-align: justify};
</style>

<div style="background-color: #FFC371; 1px; height:3px " ></div>

## 1. Introduction

**RNA-seq** is a recent approach to carry out expression profiling using high-throughput sequencing (HTS) technologies. Pre-2008, microarrays were predominantly used, but because the sequencing costs have decreased, RNA-seq became the preferred option to simultaneously measure the expression of tens of thousands of genes for multiple samples.

In this practical exercise we walk through a gene-level RNA-seq differential expression analysis using **Bioconductor** packages to find genes over- or under-expressed in bladder cancer patients.

### 1.1 Experimental data

The [**Cancer Genome Atlas**](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) (**TCGA**) is a collaboration between the National Cancer Institute (NCI) and the National Human Genome Research Institute (NHGRI) that has generated comprehensive, multi-dimensional maps of the key genomic changes in 33 types of cancer. The TCGA dataset, comprising more than two petabytes of genomic data, has been made publicly available, and this genomic information helps the cancer research community to improve the prevention, diagnosis, and treatment of cancer.

[**Recount**](https://jhubiostatistics.shinyapps.io/recount/)  is an online resource consisting of RNA-seq gene and exon counts for different studies, including TCGA data. From there, we can download the RNA-seq data of bladder cancer patients:

| **Cancer** | **Download** |
|------------|-------------------------------------------------------------------------|
| Bladder    | [Link](http://duffel.rail.bio/recount/v2/TCGA/rse_gene_bladder.Rdata)   |



The practical consist on a worked example using data from [**bladder cancer**](http://duffel.rail.bio/recount/v2/TCGA/rse_gene_bladder.Rdata), performing a differential expression analysis. 


<div style="background-color: #FFC371; 1px; height:3px " ></div>
# 2. Tools installation 

Follow `P2.Rmd` instructions if you need to install **[R](https://cran.r-project.org/)** and/or **[RStudio](https://rstudio.com/products/rstudio/download/#download)** for either Windows or Linux.

### 2.1 Required R packages

`Bioconductor` has many packages supporting analysis of high-throughput sequence data, including RNA-seq. The packages that we will use in this practical exercise include core packages maintained by the Bioconductor core team for importing and processing raw sequencing data and loading gene annotations.

```{r packages, eval = FALSE}
install.packages("ggplot2")
install.packages("ggrepel")
install.packages("BiocManager")
BiocManager::install(c("SummarizedExperiment", "DESeq2", "org.Hs.eg.db", "biomaRt", "edgeR", "tweeDEseq", "GOstats", "tweeDEseqCountData", "annotate", "biomaRt"))
install.packages("devtools") 
devtools::install_github("BlakeRMills/MetBrewer")
```

<div style="background-color: #FFC371; 1px; height:3px " ></div>

# <i class = "fa fa-search"></i>Conducting an RNA-seq analysis

With this repport, we are going to illustrate how to perform an RNA-seq expression analysis on RNA-seq data from patients diagnosed with bladder cancer. The goal is to compare the transcriptomic profile of patients in an early tumoral stage with ones in an advanced stage of this disease. The objective is to find differential expressed in this type of cancer.

The data is available in a **Ranged Summarized Experiment** (RSE) format. It is a matrix-like container where rows represent ranges of interest and columns represent samples (with sample data summarized as a `data.frame`).

### 1. Prepare data

**<i class="fa fa-cogs"></i> Create a working directory named `Project4` and create a folder named `bladderCancer`. **

```{r, eval = FALSE}
# bash code to create directories
mkdir -p Project4/bladderCancer
```

**<i class="fa fa-cogs"></i> Navigate to the `bladderlCancer` folder and download the [RSE data](http://duffel.rail.bio/recount/v2/TCGA/rse_gene_bladder.Rdata) for bladder cancer . **

<i class="fa fa-key"></i> This time you can use the `wget` function in the Linux terminal to download the file!

```{r, eval = FALSE}
# bash code to download and uncompress the data
wget "http://duffel.rail.bio/recount/v2/TCGA/rse_gene_bladder.Rdata"
```

**<i class="fa fa-cogs"></i> RSE data can be loaded into `R` with the `load()` function.**

<i class="fa fa-key"></i> Remember to change the working directory to properly load the data if it's necessary:

```{r read-RNA-seq-data}
# getwd() to know the current working directory
# setwd("~/Descargas/Project4") will change the current directory to the Project4 folder inside the Descargas folder

# Read RSE data of the cancer experiment
getwd()
load(file = "rse_gene_bladder.Rdata")
dim(rse_gene)
```

An object called `rse_gene` will be in your R environment.

Exploring the data using the `dim()` function, we see that there are `r dim(rse_gene)[1]` genes (number of rows) analyzed in a cohort of `r dim(rse_gene)[2]` patients with cancer (number of columns).

The variable `gdc_cases.diagnoses.tumor_stage` which is inside the `rse_gene` object contains information of the tumoral stage (i.e.: `r sort(unique(rse_gene$gdc_cases.diagnoses.tumor_stage))`). We can use this information to create a new variable in the `rse_gene` object called `GROUP` distinguishing "_early_" and "_late_" tumours. Early tumours are those in stages i and ii, while late tumours those in stages iii and iv. 

The following code will add an additional column `GROUP` to the meta-data to organize the cancer stages:

```{r create-group}
stage <- rse_gene$gdc_cases.diagnoses.tumor_stage

# id list of early tumours
ids.early <-grep(paste("stage i$", "stage ia$","stage ib$","stage ic$", "stage ii$", "stage iia$","stage iib$", "stage iic$",sep="|"), stage)

# id list of late tumours
ids.late <-grep(paste("stage iii$", "stage iiia$", "stage iiib$","stage iiic$", "stage iv$", "stage iva$","stage ivb$", "stage ivc$",sep="|"), stage)

# create an empty column named GROUP
colData(rse_gene)$GROUP <-rep(NA, ncol(rse_gene))
# add early for those patients with tumours at stages i-ii
colData(rse_gene)$GROUP[ids.early] <- "early"
# add late for those patients with tumours at stages iii-iv
colData(rse_gene)$GROUP[ids.late] <- "late"
```
 
<i class="fa fa-question-circle"></i> **Here we plot the number of patients in each stage using the following `data.frame()`**

```{r data-frame-fig}
# create dataframe
dataGroups <- data.frame(Stage = rse_gene$gdc_cases.diagnoses.tumor_stage, Group = rse_gene$GROUP)

# ggplot2 plot
ggplot(data = dataGroups, mapping = aes(x= Stage, fill = Group))+
  geom_bar()+
  scale_fill_manual(values=met.brewer("Derain", 3))+
  theme_bw()+
  labs(y="Number of patients")
ggsave(plot = last_plot(), file ="First_plot.png") 
```

It is important to remove all patients whose stage of the cancer was not recorded. We can check if it's necessary in the bladder cancer dataset using the `table()` function:

```{r check-na}
# Check if the status is NA
naData <- is.na(rse_gene$GROUP)
table(naData)
```

As we can see, `r table(naData)[1]` of our patients have information because their results are `FALSE` when we check if something "_is na?_". But we are dealing with `r table(naData)[2]` `TRUE` results (i.e., there's no data for some patients), we can remove it as:

```{r remove-na}
# After removing
dim(rse_gene)

# Remove NA
rse_gene <- rse_gene[, !naData]
dim(rse_gene)
```

We can summarize individuals in each category as "early" and "late" using the `table()` function. We can see that there are more than two times patients in the late stage than in the early stage.

```{r summary}
# summary of the patients
table(rse_gene$GROUP)
```

After exploring a bit the data, we can now get the **read counts** data. It can be retrieved using the `assay()` function. 

```{r save-counts}
# save the count data
counts <- assay(rse_gene, "counts")
counts[1:5, 1:2]
```

The data related to the **phenotype** of the patients can also be retrieved using the `colData()` function:

```{r save-pheno}
# save the phenotype data
phenotype <- colData(rse_gene)
phenotype[1:5, 1:2]
```
 
We need to check that the same individuals are found both in the `counts` dataset and in the `phenotype` dataset.

```{r check-ind}
# check if the same individuals are found in the datasets
identical(colnames(counts), rownames(phenotype))
```
 
Because the result is `TRUE`, it means that we have the same individuals in both datasets. If the result were `FALSE`, we would need to only keep those individuals in common in both datasets. For that, we can use the `intersect` function.

```{r keep-same-individuals}
# save a vector with the id of the individuals in common in both datasets
individualsCommon <- intersect(colnames(counts),rownames(phenotype))
# filter the count dataset to keep only the individuals in the vector
counts <- counts[, individualsCommon]
# filter the phenotype dataset to keep only the individuals in the vector
phenotype <- phenotype[individualsCommon ,]
```

In this case, we still have the `r length(individualsCommon)` individuals.

Finally, the **gene information** can be retrived using the `rowData()` function:

```{r annotation}
# save the annotation data
annotation <- rowData(rse_gene)
head(annotation)
```

We have information for a total of `r length(annotation$gene_id)` genes.

<i class="fas fa-comment-dots"></i> For each gene we have its ID, its length (in base pairs) and its symbol.

<i class="fas fa-comment-dots"></i> We have `r dim(rse_gene)[2]` individuals (differing from the pre-filtered `r length(stage)` individuals), `r dim(rse_gene)[1]` genes (with their ID, length and symbol) and there are `r table(rse_gene$GROUP)[1]` individuals in early stage and `r table(rse_gene$GROUP)[2]` individuals in late stage.

### 2. Normalization

Why is it necessary to **normalize** RNA-seq data?

1. The number of counts is related to sequencing depth

_Accounting for sequencing depth is necessary for comparison of gene expression between samples. In the example below, each gene appears to have doubled in expression in Sample A relative to Sample B, however this is a consequence of Sample A having double the sequencing depth._

<center>
  <img src = "https://hbctraining.github.io/DGE_workshop/img/normalization_methods_depth.png" width = 60%>
</center>


2. The number of counts is related to transcript length

_Accounting for gene length is necessary for comparing expression between different genes within the same sample. In the example, Gene X and Gene Y have similar levels of expression, but the number of reads mapped to Gene X would be many more than the number mapped to Gene Y because Gene X is longer._

<center>
  <img src = "https://hbctraining.github.io/DGE_workshop/img/normalization_methods_length.png" width = 30%>
</center>


3. The number of counts is proportional to the mRNA expression level

_A few highly diferentially expressed genes between samples, differences in the number of genes expressed between samples, or presence of contamination can skew some types of normalization methods. Accounting for RNA composition is recommended for accurate comparison of expression between samples, and is particularly important when performing differential expression analyses_.

<center>
  <img src = "https://hbctraining.github.io/DGE_workshop/img/normalization_methods_composition.png" width = 60%>
</center>

Information retrieved from [Introduction to DGE](https://hbctraining.github.io/DGE_workshop/lessons/02_DGE_count_normalization.html).

_The aim of normalization is to remove systematic technical effects that occur in the data to ensure that technical bias has minimal impact on the results_ ([Robinson and Oshlack, 2010](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25)).

There are different methods to normalize the counts (for a review see [Evans et al., 2017](https://academic.oup.com/bib/article/19/5/776/3056951)). We are going to focus in **two types**:

1. **RPKM** ([Mortazavi et al., 2008](https://www.nature.com/articles/nmeth.1226)) -- Reads Per Kilobase Million. This method corrects for the sequencing depth and the gene length. It's a non-sophisticated normalization method. Counts are divided by the transcript length (kb) times the total number of millions of mapped reads:

$$ RPKM = \frac{\frac{number\;of\;reads\;in\;region}{region\;length\times10^3}}{total\;reads\times10^6}$$

2. **TMM** ([Robinson and Oshlack, 2010](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25)) -- Trimmed Mean of M values. This method was developed to address this issue: _The proportion of reads attributed to a given gene in a library depends on the expression properties of the whole sample rather than just the expression level of that gene._ Therefore, the method accounts for sequencing depth, RNA composition, and gene length.

<div style = "background-color:#FFDAA1">
  <i class="fa fa-info-circle"></i> Both methods allows comparisons between genes within a sample, but only TMM is recommended for between sample comparisons and differential expression analyses.
</div>

The **MA-plots** are used to check whether normalization is needed or not. There is a function in the `edgeR` package designed to create such a plot, the `maPlot()` function:

```{r maPlot-row}
maPlot(counts[,1],
       counts[,2],
       pch=19,
       cex=.5,
       ylim=c(-8,8),
       allCol="darkgray",
       lowess=TRUE,
       xlab=expression(A == log[2] (sqrt(S1/N %.% S2/N))),
       ylab=expression(M == log[2](S1/N)-log[2](S2/N)))
grid(col="black")
title("Raw data")
```

This plot represents the  log-fold change (M-values,  i.e.  the  log  of  the  ratio  of  level  counts  for  each  gene  between two samples) against the log-average (A-values, i.e.  the average level counts for each gene across the two samples). From a MA-plot one can see if normalization is needed or not. 

One expects that the vast majority of genes are not differentially expressed between individual, thus having a symmetrical distribution on the plot with most of the genes at the 0 (`y` = 0). A lowess fit (red line) is plotted underlying a possible trend in the bias related to the mean expression.

#### 2.1 RPKM normalization

We can normalize read counts using the RPKM normalization simply applying the previous formula.

```{r rpkm}
# length of the gene
geneLength <- annotation$bp_length

# RPKM normalization method
counts.rpkm <- t(t(counts/geneLength*1000)/colSums(counts)*1e6)

counts.rpkm[1:5,1:2]
```

```{r rpkm-visualization}
maPlot(counts.rpkm[,1],
       counts.rpkm[,2],
       pch=19,
       cex=.5,
       ylim=c(-8,8),
       allCol="darkgray",
       lowess=TRUE,
       xlab=expression( A==log[2] (sqrt(S1/N%.%S2/N))),
       ylab=expression(M==log[2](S1/N)-log[2](S2/N)))
grid(col="black")
title("RPKM")
```

#### 2.2 TMM normalization

TMM normalization method is implemented in the `tweeDEseq` Bioconductor package, inside the `normalizeCounts()` function.

```{r tmm}
# TMM normalization method
counts.tmm <- normalizeCounts(counts, method = "TMM")

counts.tmm[1:5,1:2]
```

```{r tmm-visualization}
maPlot(counts.tmm[,1],
       counts.tmm[,2],
       pch=19, cex=.5,
       ylim=c(-8,8),
       allCol="darkgray",
       lowess=TRUE,
       xlab=expression( A==log[2] (sqrt(S1/N%.%S2/N))),
       ylab=expression(M==log[2](S1/N)-log[2](S2/N)))
grid(col="black")
title("TMM")
```

<center>
  <img src="http://drive.google.com/uc?export=view&id=1R7zMETMRCbOrpCYBfzi9AB08VyrTMUV7" width=80%>
</center>

<i class="fas fa-comment-dots"></i> We observe a certain deviation at the y = 0 limit between raw data and the RPKM normalization, but in the case of the TMM method, the deviation is minimal, so this one is the best normalization method.

### 3. Differential expression analysis

The R package `DESeq2` allows researchers to test differential gene expression analysis based on the negative binomial distribution ([Love, et al., 2014](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8)).

The starting point of a `DESeq2` analysis is a count matrix with one row for each gene and one column for each sample. The data is **unnormalized reads count**, because this package has its own normalization method.

This  function  requires  the  `SummarizedExperiment` object  and the design, which in this case is given in the variable `GROUP`  because we want to compare early vs. late tumor stages. 

```{r diffgen}
# stage of each patient
pheno.stage <- subset(phenotype, select=GROUP)

# recreate the counts in a new matrix
counts.adj <- matrix((as.vector(as.integer(counts))), nrow=nrow(counts), ncol=ncol(counts))

rownames(counts.adj) = rownames(counts)
colnames(counts.adj) = colnames(counts)

# check information
identical(colnames(counts.adj), rownames(pheno.stage))

# transform the group variable to factor
pheno.stage$GROUP <- as.factor(pheno.stage$GROUP)

# create the DESeqDataSet input
DEs <- DESeqDataSetFromMatrix(countData = counts.adj,
                              colData = pheno.stage,
                              design = ~ GROUP)

# differential expression analysis
dds <- DESeq(DEs)

# results extracts a result table from a DESEq analysis 
res <- results(dds, pAdjustMethod = "fdr")

head(res)
```

The `plotMA` function allows the graphical representation of the log<sub>2</sub> fold-change over the mean of normalized counts for all the samples in the `DESeqDataSet`. Points will be colored red if the adjusted p-value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down.

```{r visualization}
plotMA(res, ylim=c(-20,20), main='Differentially expressed genes in early vs late bladder cancer')
```

With this analysis it is possible to find the genes that are significanlty overexpressed or underexpressed in late tumours. We will apply the following filters:

1. Keep genes whose adjusted p-value is lower than 0.001 

```{r result1}
res.padj <- res[(res$padj < 0.001 & !is.na(res$padj)),]
nrow(res.padj)
```

We keep `r nrow(res.padj)` genes.

2. The other variable that will be useful for the filtering is the log<sub>2</sub>fold-change. We will keep genes that have a 10 log<sub>2</sub>fold-change.

```{r top}
res.padj_4fold <- res.padj[abs(res.padj$log2FoldChange) > log2(10),] 
nrow(res.padj_4fold)
```

We see that this is a more stringent criteria, because only `r nrow(res.padj_4fold)` are kept.

```{r visualize-top}
plotMA(res.padj_4fold, ylim=c(-10,10), main='Most differentially expressed genes')
```

When we represent only the most differentialy expressed genes we can see that all of them are overexpressed.


| Underexpressed | Not differentially expressed | Overexpressed |
|----------------|------------------------------|---------------|
| 0              |57662                         |48             |



### 4. Post RNA-seq analysis: visualization

In statistics, a **volcano plot** is a type of scatter-plot that is used to identify changes in large data sets. It plots significance vs. fold-change on the `y` and `x` axes, respectively. 

More specifically, a volcano plot is essentially an scatter plot, constructed by plotting the **negative log of the P-value** on the `y`-axis (usually base 10). This results in data points with low P-values (highly significant) appearing toward the top of the plot. The `x`-axis is the **log of the fold change** between the two conditions (usually base 2). Each point (gene) will be colored based on the filtering (`filter`).

```{r df}
# we first create the dataframe with the results of the DE
resDF <- as.data.frame(res)
# new column with the gene names
resDF$gene <- rownames(resDF)
# new column with TRUE/FALSE representing if a gene is deferentially expressed or not
resDF$filter <- abs(res$log2FoldChange) > log2(10) & res$padj < 0.001 
table(resDF$filter)
```

1. With the previous information, build the `ggplot()`.

```{r ggplot2}
rnaSeq <- ggplot(data = resDF, mapping = aes(x=log2FoldChange, y = -log10(padj)))+ 
  geom_point(aes(color = filter)) + 
  scale_color_manual(values=met.brewer("Derain", 3)) +
  theme_light() + 
  labs(x = "Effect size: log2FoldChange", y = "-log10(adjusted p value)") + 
  ggtitle("Volcano plot: RNA-seq analysis") +
  theme(legend.position = "none")
  
rnaSeq
```

3. One finally thing we can do is to annotate the name of the DE genes. For that we can use the `geom_label_repel()` function from the `ggrepel` library.

```{r ggplot3}

rnaSeq + geom_label_repel(data=resDF[abs(resDF$log2FoldChange) > log2(10) & resDF$padj < 0.001 ,], aes(label = as.factor(gene)), alpha = 0.75, size = 2.5, force = 1.3, max.overlaps = 50)
ggsave(last_plot(), filename = "RNA_seq.png",  width = 8, height = 6, dpi = 300)
```

### 5. Post RNA-seq analysis: enrichment analysis

A common final step to interpret gene expression data is to perform a **gene set enrichment analysis** based on the functional annotation of the differentially expressed genes. This is useful for finding out if the differentially expressed genes are associated with a certain biological process or molecular function.

The first step is to find some information of the differentially expressed genes, using the `biomaRt` package.

```{r saveGenes}
# we save a list of DE genes, and we remove the last part of the id
deGenes <- gsub("\\..*", "",rownames(res.padj_4fold))
head(deGenes)
```

The first step to  use the `biomaRt` package is to select a database and a dataset. For that we use the `useMart` function. We will use as database the `ensembl` and as a dataset the `hsapiens_gene_ensembl`.

```{r biomart}
biomart_uses <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
biomart_uses
```

The second step is to build a biomaRt query. For that we use the `getBM()` function, that needs three arguments:

1. `filter`:  define a restriction
2. `attributes`: define the values we are interested in to retrieve
3. `values`: the vector of elements we want to use for the query

```{r query}
Entrez_BM <- getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
                   filter="ensembl_gene_id",
                   values=deGenes,
                   mart=biomart_uses,
                   uniqueRows=TRUE)

dim(Entrez_BM)
head(Entrez_BM)
```

We obtain a `data.set` with two columns, one with the gene ids in the Ensembl format and another with the Entrez gene format.

<i class="fas fa-comment-dots"></i> In the new dataset there are `r dim(Entrez_BM)[1]` genes, but only 47 have information, because 3 of them have no information for the `entrezgene_id` column.

We can see that there are some NA values. We can clean them using the `na.omit` function and keep those Entrez gene ids for which we have the information

```{r clean}
Entrez_ids <- as.character(na.omit(Entrez_BM$entrezgene_id))
```

<i class="fas fa-comment-dots"></i> `r length (Entrez_ids)` genes are keeping.

The `GOstats` R package has extensive facilities for testing the association of Gene Ontology (GO) terms to genes in a gene list.

To  perform  a GO enrichment analysis, one  needs  to  define  a **gene universe** dataset and a list of the interesting genes (i.e., the DE expressed genes). 

```{r GO}
universe <- mappedkeys(org.Hs.egGO)
count.mappedkeys(org.Hs.egGO)
```

We have information for a total of `r count.mappedkeys(org.Hs.egGO)` human genes.

```{r testGO}
# Set the threshold for the p value to 5%
GOtest <- new("GOHyperGParams",
                 geneIds = Entrez_ids,
                 universeGeneIds=universe,
                 annotation = "org.Hs.eg.db",
                 ontology = "BP",
                 pvalueCutoff= 0.05,
                 conditional = FALSE,
                 testDirection = "over") 
# hypergeometric test
GOtestOver <- hyperGTest(GOtest)
GOtestOver

GOresult <- summary(GOtestOver)
head(GOresult)

```

```{r data-frame-fig2}
# ggplot2 plot
filter <- GOresult[GOresult$Count >4, ]
filter

GOplot <- ggplot(data = filter, mapping = aes(x = Term, y = Count, fill = Term)) +
  geom_col(show.legend = FALSE, width = 0.75) +
  theme(axis.text.x = element_text(angle = 90))+
  scale_fill_manual(values=met.brewer("Derain", 33))
  labs(title = "Gene Ontology Enrichment Analysis", x = "GO Term")
GOplot

ggsave(last_plot(), filename = "GO.png",  width = 8, height = 6, dpi = 300)
```

<i class="fas fa-comment-dots"></i> We can observe that the most represented GO term is the tissue development, but without many differences with respect to the other GO terms in the chart.


<div style="background-color: #FFC371; 1px; height:3px " ></div>

# 5. References

Practical based on Juan Ramón González' material available at [GitHub](https://github.com/isglobal-brge/TeachingMaterials/tree/master/Master_Bioinformatics).

<div style="background-color: #FFC371; 1px; height:3px " ></div>
<br>